---
title: "Spatial robustness of patch-level FPCA1"
subtitle: "dbMEM variance partitioning and SPDE mixed-model sensitivity analysis"
author: "Derek Corcoran and collaborators"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  bookdown::pdf_document2:
    toc: true
    number_sections: true
    fig_caption: true
    keep_tex: true
    latex_engine: pdflatex
bibliography: spatial_references.bib
link-citations: true
geometry: margin=1in
fontsize: 10pt
params:
  project_root: "."
  fpca_results_dir: "Results"
  patch_dir: "Species_PatchDistances"
  spatial_output_dir: "Results/spatial_autocorrelation"
  merge_variant: "20m"
  analysis_crs: 25832
  mem_grid_km: 10
  mem_method: "MIR"
  mem_nperm: 999
  mem_nperm_global: 9999
  mesh_cutoffs_km:
    value: [10, 20, 40]
  primary_cutoff_km: 20
  residual_grid_km:
    value: [2.5, 5, 10]
  residual_nsim: 500
  run_centroid_sensitivity: true
  seed: 37
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 7.2,
  fig.height = 5.2,
  dpi = 300,
  cache = TRUE,
  collapse = TRUE,
  comment = "#>"
)
options(stringsAsFactors = FALSE, scipen = 999)

input_path <- knitr::current_input(dir = TRUE)
analysis_dir <- if (dir.exists(input_path)) input_path else dirname(input_path)
source(file.path(analysis_dir, "spatial_analysis_helpers.R"))
assert_spatial_packages()

resolve_path <- function(path, root) {
  is_absolute <- grepl("^(/|[A-Za-z]:[/\\\\])", path)
  normalizePath(
    if (is_absolute) path else file.path(root, path),
    winslash = "/",
    mustWork = FALSE
  )
}

project_root <- normalizePath(params$project_root, winslash = "/", mustWork = TRUE)
fpca_results_dir <- resolve_path(params$fpca_results_dir, project_root)
patch_dir <- resolve_path(params$patch_dir, project_root)
spatial_output_dir <- resolve_path(params$spatial_output_dir, project_root)
dir.create(spatial_output_dir, recursive = TRUE, showWarnings = FALSE)

mesh_cutoffs_km <- as.numeric(unlist(params$mesh_cutoffs_km))
residual_grid_km <- as.numeric(unlist(params$residual_grid_km))
stopifnot(params$primary_cutoff_km %in% mesh_cutoffs_km)
set.seed(params$seed)
```

# Purpose and inferential architecture

This report replaces linear coordinate terms ($x+y+xy$) with two complementary
representations of spatial dependence. First, Moran eigenvector maps (dbMEMs)
represent residual spatial structure at multiple scales in the variance
partition. Second, a continuous Matérn spatial random field is fitted with the
SPDE approach in `sdmTMB` [@lindgren2011; @anderson_sdmTMB]. The same fixed and
species-random-effect structure is fitted with and without the spatial field.

The analysis deliberately keeps two geometries separate:

1. **Connectivity geometry:** FPCA1 is derived upstream from ordered,
   border-to-border distances among habitat polygons.
2. **Residual geographic structure:** dbMEM and SPDE use representative
   geographic locations of the final patches, never the edge-distance matrix
   that generated FPCA1.

This separation avoids using the response-generating edge distances again as
spatial controls. Edge-to-edge distances remain the ecologically appropriate
connectivity measure because centroid distances can substantially overstate the
physical gap between large or elongated patches.

This report assumes that the upstream ordered profiles are the intended ones.
The current data-generation workflow preselects up to 15 candidate neighbours
by centroid distance and then calculates and ranks polygon edge distances within
that candidate set. The resulting values are edge-to-edge distances, but the
candidate shortcut does not mathematically guarantee that they are the five
globally nearest polygon edges for large or elongated patches. Before the final
manuscript, a targeted exact-neighbour audit of the largest and least compact
patches is therefore advisable.

The selected patch correction is **`r spatial_variant_spec(params$merge_variant)$label`**.
The report stops if the FPCA identifiers do not match the geometry, lookup, and
area files for that exact correction.

# Input assembly and validation

```{r read-and-join-data}
fpca_scores <- read_fpca_patch_scores(fpca_results_dir)

analysis_sf <- assemble_spatial_analysis_data(
  scores = fpca_scores,
  patch_dir = patch_dir,
  merge_variant = params$merge_variant,
  analysis_crs = params$analysis_crs,
  strict = TRUE
)

model_data <- prepare_model_data(analysis_sf)

saveRDS(
  list(
    model_data = model_data,
    parameters = params,
    generated = Sys.time()
  ),
  file.path(spatial_output_dir, "spatial_model_data.rds")
)

scenario_counts <- model_data |>
  dplyr::count(.data$species, .data$mapping, .data$quality, name = "n_patches") |>
  dplyr::arrange(.data$mapping, .data$quality, .data$species)

readr::write_csv(
  scenario_counts,
  file.path(spatial_output_dir, "input_patch_counts.csv")
)

knitr::kable(
  scenario_counts,
  caption = "Patch-level observations entering the spatial analyses.",
  booktabs = TRUE,
  longtable = TRUE
)
```

FPCA signs are mathematically arbitrary. The import function orients each
scenario so that larger FPCA1 values represent shorter ordered distances and
therefore greater local structural connectivity. The response and log patch
area are standardized globally for model fitting; they are not standardized
within habitat definition, because within-definition standardization would
erase the mapping and quality contrasts of interest.

## Polygon-support diagnostics

A point on surface is used for the primary spatial representation. It is
guaranteed to lie in the dissolved final patch, but it remains a one-point
approximation. The geometric centroid is therefore retained as an explicit
sensitivity analysis rather than treated as a perfect description of a large
or complex polygon [@pebesma2018].

```{r support-summary}
support_summary <- geometry_support_summary(model_data)
readr::write_csv(
  support_summary,
  file.path(spatial_output_dir, "polygon_support_summary.csv")
)
knitr::kable(
  support_summary,
  digits = 3,
  caption = "Distribution of polygon size, compactness, and displacement between centroid and point-on-surface representations.",
  booktabs = TRUE
)
```

```{r largest-patches, fig.cap="The nine largest final patches, with point-on-surface (blue) and geometric centroid (red) representations."}
largest_patches <- analysis_sf |>
  dplyr::arrange(dplyr::desc(.data$area_km2)) |>
  dplyr::slice_head(n = 9L)

largest_surface <- suppressWarnings(sf::st_point_on_surface(largest_patches))
largest_centroid <- suppressWarnings(sf::st_centroid(largest_patches))

ggplot2::ggplot() +
  ggplot2::geom_sf(data = largest_patches, fill = "grey90", colour = "grey35") +
  ggplot2::geom_sf(data = largest_surface, colour = "#2166AC", size = 1.8) +
  ggplot2::geom_sf(data = largest_centroid, colour = "#B2182B", shape = 4, size = 1.8) +
  ggplot2::facet_wrap(~ species + scenario, scales = "free") +
  ggplot2::theme_bw(base_size = 8) +
  ggplot2::theme(axis.text = ggplot2::element_blank(), axis.ticks = ggplot2::element_blank()) +
  ggplot2::labs(x = NULL, y = NULL)
```

# dbMEM variance partitioning

Full point-level dbMEM eigendecomposition scales poorly with the square of the
number of locations and may be infeasible for these patch sets. We therefore
calculate dbMEMs on occupied **`r params$mem_grid_km` km** cells and map the
selected cell-level eigenvectors back to the patch observations. This is a
reduced-rank spatial block: it represents broad-to-intermediate residual
spatial structure while preserving patch-level ecological predictors.

MEM selection is applied to cell-mean residuals from

$$
\mathrm{FPCA1} \sim \log(\mathrm{area}) + \mathrm{species} +
\mathrm{mapping} + \mathrm{quality},
$$

not to raw FPCA1. The **`r params$mem_method`** criterion is used; MIR adds
eigenvectors until residual Moran's $I$ is no longer significant
[@bauman2018; @dray2006]. This prevents the spatial block from being selected
simply because ecological predictors themselves are geographically structured.
Because eigenvectors are response-selected, the resulting variance fractions
are treated as descriptive effect partitioning rather than independent
hypothesis tests.

```{r select-mems}
mem_point <- select_mem_block(
  data = model_data,
  xy_cols = c("point_X_km", "point_Y_km"),
  grid_km = params$mem_grid_km,
  method = params$mem_method,
  nperm = params$mem_nperm,
  nperm_global = params$mem_nperm_global
)

mem_centroid <- select_mem_block(
  data = model_data,
  xy_cols = c("centroid_X_km", "centroid_Y_km"),
  grid_km = params$mem_grid_km,
  method = params$mem_method,
  nperm = params$mem_nperm,
  nperm_global = params$mem_nperm_global
)

mem_summary <- dplyr::bind_rows(
  data.frame(
    support = "point_on_surface",
    occupied_cells = nrow(mem_point$cells),
    positive_MEMs = mem_point$all_mem_count,
    selected_MEMs = mem_point$selected_mem_count
  ),
  data.frame(
    support = "centroid",
    occupied_cells = nrow(mem_centroid$cells),
    positive_MEMs = mem_centroid$all_mem_count,
    selected_MEMs = mem_centroid$selected_mem_count
  )
)

readr::write_csv(mem_summary, file.path(spatial_output_dir, "mem_selection_summary.csv"))
knitr::kable(
  mem_summary,
  caption = "dbMEM construction and selection under two spatial-support representations.",
  booktabs = TRUE
)
```

## Four-block partition

The response is partitioned among patch area, species identity, habitat
definition (mapping and quality), and selected MEMs. With a univariate
response, `vegan::varpart()` performs partial linear regression and reports
adjusted $R^2$ fractions [@peresneto2006; @oksanen_vegan].

```{r variance-partition}
vp_point <- run_variance_partition(model_data, mem_point)
vp_centroid <- run_variance_partition(model_data, mem_centroid)

partition_comparison <- dplyr::bind_rows(
  dplyr::mutate(vp_point$table, support = "point_on_surface"),
  dplyr::mutate(vp_centroid$table, support = "centroid")
) |>
  dplyr::select(
    .data$support, .data$block, .data$marginal_adjusted_r2,
    .data$unique_adjusted_r2, .data$full_model_adjusted_r2
  )

readr::write_csv(
  partition_comparison,
  file.path(spatial_output_dir, "variance_partition_support_sensitivity.csv")
)

partition_fractions <- dplyr::bind_rows(
  varpart_fraction_table(vp_point$varpart, "point_on_surface"),
  varpart_fraction_table(vp_centroid$varpart, "centroid")
)
readr::write_csv(
  partition_fractions,
  file.path(spatial_output_dir, "variance_partition_all_fractions.csv")
)

knitr::kable(
  partition_comparison,
  digits = 4,
  caption = "Adjusted R-squared associations and unique fractions under point-on-surface and centroid spatial support.",
  booktabs = TRUE
)
```

```{r variance-partition-fractions}
knitr::kable(
  partition_fractions,
  digits = 4,
  caption = "Complete testable and shared fractions from vegan::varpart under both spatial-support representations.",
  booktabs = TRUE,
  longtable = TRUE
)
```

```{r variance-partition-plot, fig.cap="Variation partitioning with point-on-surface dbMEMs. Fractions can be negative after adjusted-R-squared correction and should not be truncated when reported."}
graphics::par(mar = c(1, 1, 2, 1))
plot(
  vp_point$varpart,
  Xnames = names(vp_point$blocks),
  bg = c("#4C78A8", "#F58518", "#54A24B", "#B279A2")[seq_along(vp_point$blocks)]
)
```

# SPDE mixed models

The primary model comparison holds the ecological model constant:

$$
M_0:\quad \mathrm{FPCA1}_i = \beta_0 + \beta_1\log(\mathrm{area}_i) +
\mathrm{mapping}_i + \mathrm{quality}_i + u_{\mathrm{species}(i)} + \epsilon_i,
$$

$$
M_S:\quad \mathrm{FPCA1}_i = \beta_0 + \beta_1\log(\mathrm{area}_i) +
\mathrm{mapping}_i + \mathrm{quality}_i + u_{\mathrm{species}(i)} +
\omega(s_i) + \epsilon_i,
$$

where $\omega(s_i)$ is a continuous Matérn spatial field. `sdmTMB` supports
both the SPDE field and lme4-style species random intercepts
[@anderson_sdmTMB].

```{r fit-spde-models}
spde_point <- fit_spde_suite(
  data = model_data,
  xy_cols = c("point_X_km", "point_Y_km"),
  cutoffs_km = mesh_cutoffs_km
)
spde_point_tidy <- tidy_spde_suite(spde_point)
effect_comparison <- compare_fixed_effects(
  spde_point_tidy$fixed,
  primary_cutoff_km = params$primary_cutoff_km
)

saveRDS(spde_point, file.path(spatial_output_dir, "spde_point_on_surface_models.rds"))
readr::write_csv(spde_point_tidy$fixed, file.path(spatial_output_dir, "spde_fixed_effects.csv"))
readr::write_csv(spde_point_tidy$random_parameters, file.path(spatial_output_dir, "spde_random_parameters.csv"))
readr::write_csv(spde_point_tidy$aic, file.path(spatial_output_dir, "spde_model_aic.csv"))
readr::write_csv(effect_comparison, file.path(spatial_output_dir, "spde_effect_change.csv"))
```

## Mesh sensitivity and convergence

A cutoff is the minimum distance between candidate mesh vertices in the
coordinate units (kilometres here). Smaller cutoffs create finer, slower
meshes. Conclusions are evaluated across **`r paste(mesh_cutoffs_km, collapse = ", ")` km**
instead of relying on one arbitrary resolution.

```{r mesh-summary-table}
knitr::kable(
  spde_point_tidy$aic,
  digits = 2,
  caption = "Non-spatial and SPDE model comparison across mesh cutoffs.",
  booktabs = TRUE
)
```

```{r primary-mesh, fig.cap="Primary SPDE mesh using point-on-surface patch locations."}
primary_name <- paste0("cutoff_", params$primary_cutoff_km, "km")
plot(spde_point$meshes[[primary_name]])
```

```{r model-sanity}
primary_spatial_model <- spde_point$spatial[[primary_name]]
sanity_output <- utils::capture.output(sdmTMB::sanity(primary_spatial_model))
cat(paste(sanity_output, collapse = "\n"))
```

## Do ecological effects persist?

```{r fixed-effect-table}
knitr::kable(
  effect_comparison,
  digits = 4,
  caption = paste0(
    "Fixed effects before and after adding the primary ",
    params$primary_cutoff_km, " km SPDE field."
  ),
  booktabs = TRUE
)
```

```{r fixed-effect-plot, fig.cap="Fixed-effect estimates across the non-spatial model and SPDE mesh resolutions. Error bars are 95% confidence intervals."}
effect_plot_data <- spde_point_tidy$fixed |>
  dplyr::filter(.data$term != "(Intercept)") |>
  dplyr::mutate(
    model_label = dplyr::if_else(
      .data$model == "nonspatial",
      "Non-spatial",
      paste0("SPDE ", .data$cutoff_km, " km")
    )
  )

ggplot2::ggplot(
  effect_plot_data,
  ggplot2::aes(x = .data$estimate, y = .data$term, colour = .data$model_label)
) +
  ggplot2::geom_vline(xintercept = 0, colour = "grey70") +
  ggplot2::geom_errorbarh(
    ggplot2::aes(xmin = .data$conf.low, xmax = .data$conf.high),
    height = 0.15,
    position = ggplot2::position_dodge(width = 0.55)
  ) +
  ggplot2::geom_point(position = ggplot2::position_dodge(width = 0.55), size = 1.8) +
  ggplot2::theme_bw() +
  ggplot2::labs(x = "Standardized coefficient", y = NULL, colour = "Model")
```

```{r spatial-parameter-table}
knitr::kable(
  spde_point_tidy$random_parameters,
  digits = 4,
  caption = "Estimated random-effect and Matérn parameters across mesh resolutions. Range is in kilometres.",
  booktabs = TRUE
)
```

## Point-on-surface versus centroid SPDE

```{r centroid-spde-sensitivity}
if (isTRUE(params$run_centroid_sensitivity)) {
  spde_centroid <- fit_spde_suite(
    data = model_data,
    xy_cols = c("centroid_X_km", "centroid_Y_km"),
    cutoffs_km = params$primary_cutoff_km
  )
  centroid_tidy <- tidy_spde_suite(spde_centroid)
  support_fixed <- dplyr::bind_rows(
    spde_point_tidy$fixed |>
      dplyr::filter(
        .data$model == "SPDE",
        .data$cutoff_km == params$primary_cutoff_km
      ) |>
      dplyr::mutate(support = "point_on_surface"),
    centroid_tidy$fixed |>
      dplyr::filter(.data$model == "SPDE") |>
      dplyr::mutate(support = "centroid")
  )
  saveRDS(spde_centroid, file.path(spatial_output_dir, "spde_centroid_model.rds"))
  readr::write_csv(
    support_fixed,
    file.path(spatial_output_dir, "spde_support_sensitivity.csv")
  )
  knitr::kable(
    support_fixed,
    digits = 4,
    caption = "Fixed-effect sensitivity to point-on-surface versus centroid spatial support.",
    booktabs = TRUE
  )
}
```

## Estimated residual spatial field

```{r spatial-field-map, fig.cap="Estimated persistent spatial-field deviations, averaged within 5 km cells for display."}
field_grid <- spatial_field_at_observations(
  primary_spatial_model,
  model_data,
  grid_km = 5
)

ggplot2::ggplot(
  field_grid,
  ggplot2::aes(x = .data$grid_x, y = .data$grid_y, fill = .data$omega_s)
) +
  ggplot2::geom_tile() +
  ggplot2::coord_equal() +
  ggplot2::scale_fill_gradient2(
    low = "#2166AC", mid = "white", high = "#B2182B", midpoint = 0
  ) +
  ggplot2::theme_bw() +
  ggplot2::labs(x = "Easting (km)", y = "Northing (km)", fill = expression(omega(s)))
```

# Residual spatial autocorrelation

DHARMa requires unique coordinates for its distance-based Moran test. Because
species and habitat definitions can repeat locations, residuals are aggregated
to spatial cells before testing, as recommended in the DHARMa documentation
[@hartig_DHARMa]. Results are shown at several cell sizes to avoid depending on
one arbitrary grouping scale. Simulations use `type = "mle-mvn"`, following
the current `sdmTMB` residual workflow for random-effect models.

```{r dharma-residuals}
dharma_nonspatial <- make_dharma_residuals(
  spde_point$nonspatial,
  nsim = params$residual_nsim,
  seed = params$seed
)
dharma_spatial <- make_dharma_residuals(
  primary_spatial_model,
  nsim = params$residual_nsim,
  seed = params$seed
)

residual_tests <- dplyr::bind_rows(lapply(residual_grid_km, function(grid_km) {
  nonspatial_test <- dharma_spatial_test_from_residuals(
    residuals = dharma_nonspatial,
    data = model_data,
    xy_cols = c("point_X_km", "point_Y_km"),
    grid_km = grid_km,
    seed = params$seed,
    nsim = params$residual_nsim
  )
  spatial_test <- dharma_spatial_test_from_residuals(
    residuals = dharma_spatial,
    data = model_data,
    xy_cols = c("point_X_km", "point_Y_km"),
    grid_km = grid_km,
    seed = params$seed,
    nsim = params$residual_nsim
  )
  dplyr::bind_rows(
    dplyr::mutate(nonspatial_test$table, model = "nonspatial"),
    dplyr::mutate(spatial_test$table, model = "SPDE")
  )
})) |>
  dplyr::select(
    .data$model, .data$grid_km, .data$n_spatial_cells,
    .data$moran_observed, .data$moran_expected, .data$p_value, .data$nsim
  )

readr::write_csv(
  residual_tests,
  file.path(spatial_output_dir, "dharma_spatial_autocorrelation.csv")
)
knitr::kable(
  residual_tests,
  digits = 5,
  caption = "Distance-based Moran tests of grouped DHARMa residuals before and after adding the SPDE field.",
  booktabs = TRUE
)
```

# Interpretation and manuscript update

The evidential question is not whether a linear $x/y$ trend is small. It is
whether ecological effect estimates are stable after explicitly modelling
residual spatial dependence, and whether meaningful residual autocorrelation
remains.

```{r automatic-interpretation, results='asis'}
area_effect <- effect_comparison |>
  dplyr::filter(.data$term == "log_area_z")

primary_range <- spde_point_tidy$random_parameters |>
  dplyr::filter(
    .data$cutoff_km == params$primary_cutoff_km,
    .data$term == "range"
  ) |>
  dplyr::slice_head(n = 1L)

primary_sigma <- spde_point_tidy$random_parameters |>
  dplyr::filter(
    .data$cutoff_km == params$primary_cutoff_km,
    .data$term == "sigma_O"
  ) |>
  dplyr::slice_head(n = 1L)

cat("The primary point-on-surface SPDE model estimated a Matérn range of ")
cat(if (nrow(primary_range)) sprintf("%.2f km", primary_range$estimate) else "NA")
cat(" and a spatial-field standard deviation of ")
cat(if (nrow(primary_sigma)) sprintf("%.3f", primary_sigma$estimate) else "NA")
cat(" on the standardized FPCA1 scale. ")

if (nrow(area_effect)) {
  cat(sprintf(
    "The standardized log-area coefficient changed from %.3f (95%% CI %.3f to %.3f) without the field to %.3f (95%% CI %.3f to %.3f) with the field. ",
    area_effect$nonspatial_estimate,
    area_effect$nonspatial_low,
    area_effect$nonspatial_high,
    area_effect$spatial_estimate,
    area_effect$spatial_low,
    area_effect$spatial_high
  ))
}

primary_residual <- residual_tests |>
  dplyr::filter(.data$grid_km == residual_grid_km[[which.min(abs(
    residual_grid_km - 5
  ))]]) |>
  dplyr::arrange(.data$model)

if (nrow(primary_residual) == 2L) {
  non_row <- primary_residual[primary_residual$model == "nonspatial", ]
  sp_row <- primary_residual[primary_residual$model == "SPDE", ]
  cat(sprintf(
    "At the %.1f km residual grouping scale, Moran's I changed from %.4f (p = %.4g) to %.4f (p = %.4g) after adding the field.",
    non_row$grid_km,
    non_row$moran_observed,
    non_row$p_value,
    sp_row$moran_observed,
    sp_row$p_value
  ))
}
```

If the area, mapping, and quality coefficients retain their direction and
similar magnitude across mesh resolutions and spatial-support choices, the
appropriate conclusion is:

> After accounting for residual spatial dependence with a continuous Matérn
> field, the estimated effects of patch size and habitat definition were
> largely unchanged. Residual distance-based Moran tests showed [limited / no]
> remaining spatial autocorrelation across the tested aggregation scales.

If one or more effects collapse or reverse, they should instead be described as
partly confounded with geographically structured variation. In either case,
the earlier statement that FPCA1 was "largely independent of geographic
location" should be removed. The old $x/y$ model supports only the narrower
claim that linear broad-scale geographic trends were weak.

## Methods-ready wording

> Edge-to-edge rather than centroid-to-centroid distances were used to derive
> ordered nearest-neighbour profiles because habitat patches varied in size and
> geometry, and centroid distances can substantially overestimate the physical
> separation of neighbouring patches. Residual spatial structure was evaluated
> independently of those edge distances. For variance partitioning, positive
> Moran eigenvector maps were constructed from occupied
> `r params$mem_grid_km`-km cells defined by patch points on surface. A subset
> was selected from residuals of the non-spatial ecological model using the
> `r params$mem_method` criterion. We partitioned adjusted $R^2$ in FPCA1 among
> patch area, species identity, habitat definition, and the selected spatial
> eigenvectors. Separately, we fitted otherwise identical Gaussian mixed models
> without and with a continuous Matérn spatial random field using the SPDE
> implementation in `sdmTMB`, with species as a random intercept. Mesh cutoffs
> of `r paste(mesh_cutoffs_km, collapse = ", ")` km were compared, and centroid
> coordinates were used as a spatial-support sensitivity analysis. Residual
> spatial autocorrelation was assessed with distance-based Moran tests on
> simulation-based DHARMa residuals aggregated to unique spatial cells.

# Reproducibility record

```{r reproducibility}
run_manifest <- data.frame(
  parameter = c(
    "merge_variant", "analysis_crs", "mem_grid_km", "mem_method",
    "mesh_cutoffs_km", "primary_cutoff_km", "residual_grid_km",
    "residual_nsim", "seed"
  ),
  value = c(
    params$merge_variant,
    params$analysis_crs,
    params$mem_grid_km,
    params$mem_method,
    paste(mesh_cutoffs_km, collapse = ","),
    params$primary_cutoff_km,
    paste(residual_grid_km, collapse = ","),
    params$residual_nsim,
    params$seed
  )
)
readr::write_csv(run_manifest, file.path(spatial_output_dir, "run_manifest.csv"))
knitr::kable(run_manifest, caption = "Analysis parameters.", booktabs = TRUE)
```

```{r session-info}
sessionInfo()
```

# References
